%% ============================
%%
%% Appendix A
%%
%% ============================

\chapter{General Relativity \& Numerical relativity}
\label{app:gr}
\externaldocument{intro}
%% --------------------------------------

\section{General-relativistic hydrodynamics}

%% --------------------------------------


This section is meant to sketch several important parts of the mathematical background. 
We focus on the aspects relevant for the tools and methods employed in out discussion. 
We do not aim to provide a comprehensive overview. 
The chapter is divided into 
\todo{list the parts and their content}

%% [ GENERAL RELATIVITY from EFE to CCZ4 ]

%% ---

\subsection{The Cauchy Problem in General Relativity}

%% ---

In this section we briefly recall the initial-value formulation of the Einstein equations of general relativity through the following steps. 
We start by introducing notations and the basics of GR. 
We summarize the Einstein field equations. 
Then we continue with how EFE can be split in a set of evolutionary equations and constraints. 
For that we focus on the Arnowitt, Deser and Misner, or ADM, formalism. 
In the end we comment on the stability of the ADM equations, on the need for strongly-hyperbolic formulations of the EFE, and on the choice of gauge conditions commonly used to evolve spacetimes with singularities. 
This overview is based in \cite{Arnowitt:1962hi,Landau:1982dva,Wald:1984,Misner:1973,Baumgarte:2002jm}, which we refer to for more detained discussion.

%% --- 
\subsubsection{Euler-Lagrange equations}
%% ---

\red{Requires understanding of: vectors, differential forms and tensors;
    differential equation types: ellipitc, parabolic, hyperbolic, their nmerical advantages;
    smooth manifolds, Lorenzian maniforld and metric, affine connection (Levi-Civita connection), covariant derivatives, action pinciple, lagrange field theory, variation of the action, Einstein-Hilbert action, Ricci scalar and Ricci tensor, Riemann tensor;
    spacelike, timelike and null hypersurface, hyperbolic space-time, space-time foliation,
    Lie-derivative along the vector field, Lagrangian density, Legendre transformation, Hamiltonian,
    extrinsic curvature, Codazzi equations, Gauss equations,
    palantini-type vatiation
    gauge conditions: slicing and spatial,}

We consider a spacetime defined by the real \textit{smooth manifold} $\mathcal{M}$ and \textit{Lorentzian metric} $\boldsymbol{g}$ on $\mathcal{M}$ of signature $(-,+,+,+)$. 
The $\nabla$ denotes the \textit{affine connection} associated with $\boldsymbol{g}$, the Levi-Civita connection.
We use the convention that all Greek indices lie in $\{0, 1, 2, 3\}$ and Lower case Latin indices $\{1, 2, 3\}$.
The $\nabla\boldsymbol{T}$ denotes the \textit{covariant derivative} of a tensor $\boldsymbol{T}$ and $\nabla_{\boldsymbol{u}}\boldsymbol{T}$ -- covariant derivative along a given vector field $\boldsymbol{u}$.

The scalar product of two vectors then 

\begin{equation}
\boldsymbol{a}\cdot\boldsymbol{b}:=g_{\mu\nu}a^{\mu}b^{\nu}
\end{equation}

The action of a linear form on a vector however is represented as 

\begin{equation}
\langle\boldsymbol{\omega},\boldsymbol{\upsilon}\rangle=\omega_{\mu}\upsilon^{\mu}
\end{equation}

Let the $\boldsymbol{\alpha}$ be the \textit{totally antisymmetric symbol} that expresses through coordinates $x^{\mu}$ as

\begin{equation}
\boldsymbol{\alpha} = dx^0 \wedge dx^1 \wedge dx^2 \wedge dx^3,
\end{equation}

where $\wedge$ denotes \textit{exterior product}. 
Then, proper \textit{volume pseudo-form} of the spacetime is

\begin{equation}
\boldsymbol{\varepsilon} = \sqrt{-g}\boldsymbol{\alpha},
\end{equation}

where $g$ denotes the determinant of the spacetime metric.

In GR, the spacetime is represented by \textit{Lorentzian manifold} $\mathcal{M}$ and $g$, the \textit{Lorentzian metric}.

The \textit{action principle} of the \textit{Lagrangian field theory} on the spacetime $(\mathcal{M}; \boldsymbol{g})$ is

\begin{equation}
S(\boldsymbol{q}, \nabla\boldsymbol{q}) = \int_{\mathcal{M}}\boldsymbol{\alpha}\mathcal{L}(\boldsymbol{q}, \nabla\boldsymbol{q}),
\end{equation}

where $\boldsymbol{q}$ are a set of generalized coordinates for the fields described by the theory, $\nabla$ is the Levi-Civita connection, $\mathcal{L}$ is a scalar density of a scalar quantity $\lambda$ as $\lambda(\boldsymbol{q},\nabla\boldsymbol{q})$. 

Varying the action with respect to the $\boldsymbol{q}$

\begin{equation}
\delta S(\boldsymbol{q}, \nabla\boldsymbol{q}) = \delta\int\boldsymbol{\alpha}\mathcal{L}(\boldsymbol{q}, \nabla\boldsymbol{q}) = \int\boldsymbol{\alpha}\Big(\frac{\partial\mathcal{L}}{\partial\boldsymbol{q}}\delta\boldsymbol{q}+\frac{\partial\mathcal{L}}{\partial(\nabla\boldsymbol{q})}\delta\nabla\boldsymbol{q}\Big)
\end{equation}

As $\delta$ and $\nabla$ commute, and partially integrating $\nabla$, we obtain

\begin{equation}
\partial S(\boldsymbol{q}, \nabla\boldsymbol{q}) = \int\boldsymbol{\alpha}\Big(\frac{\mathcal{L}}{\partial\boldsymbol{q}}-\nabla\frac{\partial \mathcal{L}}{\partial(\nabla\boldsymbol{q})}\Big)\delta\boldsymbol{q} + \int_{\mathcal{M}}\boldsymbol{\alpha}\nabla\Big(\frac{\partial\mathcal{L}}{\partial(\nabla\boldsymbol{q})}\delta\boldsymbol{q}\Big)
\end{equation}

The last term is a boundary term and in order to vanish we impose boundary condition. 
Assume that the fields are defined over only a \textit{compact domain}. \\
As the choice of $\partial\boldsymbol{q}$ is arbitrary, the 

\begin{equation}
\partial S(\boldsymbol{q}, \nabla\boldsymbol{q}) = 0
\end{equation}

and the \textit{Euler-Lagrange equations} are

\begin{equation}
\frac{\partial \mathcal{L}}{\partial\boldsymbol{q}} - \nabla\Big(\frac{\partial\mathcal{L}}{\partial(\nabla\boldsymbol{q})}\Big) = 0
\label{eq:theory:eulerlagrange}
\end{equation}

%% ---
\subsubsection{The Hilbert Action}
%% ---

The \textit{Einstein-Hilbert action} allows to obtain an \textit{Einstein field equations} through the \textit{principle of least action}. Here we briefly underline the procedure.

Introduce action that describes the gravitational field, and a matter field $\mathcal{L}_m$:

\begin{align}
S_g &= \int\frac{1}{2\kappa}R\epsilon, \\
S_m &= \int\mathcal{L}_{m}\epsilon,
\end{align}

where $R$ is the Ricci scalar and $\kappa$ is the  Einstein's constant.

The full action then:

\begin{equation}
S = \int\Big(\frac{1}{2\kappa}R+\mathcal{L}_m\Big)\epsilon
\end{equation}

The action principle dicatates, that $\delta S = 0$  with respect to the inverse metric $g^{\mu\nu}$. 

\begin{equation}
\int\Bigg[\frac{1}{2\kappa}\Big(\frac{\delta R}{\delta g^{\mu\nu}}+\frac{R}{\sqrt{-g}}\frac{\delta\sqrt{-g}}{\delta g^{\mu\nu}}\Big) + \frac{1}{\sqrt{-g}}\frac{\delta(\sqrt{-g}\mathcal{L}_m)}{\delta g^{\mu\nu}}\Bigg]\delta g^{\mu\nu}\epsilon
\end{equation}

Owing to the arbitrariness of $\delta g^{\mu\nu}$, the integrand must be zero. 

\begin{equation}
\frac{\delta R}{\delta g^{\mu\nu}} + \frac{R}{\sqrt{-g}}\frac{\delta\sqrt{-g}}{\delta g^{\mu\nu}} = -2\kappa\frac{1}{\sqrt{-g}}\frac{\delta(\sqrt{-g}\mathcal{L}_m)}{\delta g^{\mu\nu}} = -\frac{2\kappa}{\sqrt{-g}}\frac{\delta S_m}{\delta g_{\mu\nu}} := \kappa T_{\mu\nu},
\label{eq:theory:action1}
\end{equation}

where we introduced the stress-energy tensor $T_{\mu\nu}$ and the matter action $S_m$ for future use. \\

\gray{this matter action is used in deriving the $T_{\mu} ^{\nu}$ i the invariant fluid formalism}

The continuation of this derivation requires taking variation of the Riccia scalar $R$ and the determinant of the metric $\sqrt{-g}$. 
As this is a length procedure, we provide here the result. 

\begin{equation}
\frac{\delta R}{\delta g^{\mu\nu}} = R_{\mu\nu},
\label{eq:theory:deltaR}
\end{equation}

where the $R_{\mu\nu}$ is the Ricci curvature tensor.

\begin{equation}
\frac{1}{\sqrt{-g}}\frac{\delta\sqrt{-g}}{\delta g^{\mu\nu}} = -\frac{1}{2}g_{\mu\nu}.
\label{eq:theory:deltagmuny}
\end{equation}

Substituting Eq. \ref{eq:theory:deltaR} and Eq. \ref{eq:theory:deltagmuny} into equation of motion Eq.  \ref{eq:theory:action1} we obtain the Einstein's field equation 

\begin{equation}
R_{\mu\nu} -\frac{1}{2}g_{\mu\nu}R=8\pi T_{\mu\nu},
\label{eq:theory:EFE}
\end{equation}

where in the geometrized unit system, \textit{i.e} $c=G=1$, the $\kappa=8\pi$.

%% ---
\subsubsection{3+1 Decomposition of Einstein field equations}
%% ---

The Einstein field equations (\ref{eq:theory:EFE}) represent a set of $10$ non-linear partial differential equations.
These equations can be defined on a while metric $\mathcal{M}$ or a domain $\Omega\subset\mathcal{M}$, where in the latter, the boundary conditions on $\partial\Omega$ are required. 

It is convenient to chose a \textit{null hyersurface} $\Sigma\subset\mathcal{M}$ on which to define the initial data, from which the evolution of space-time begins. 
This, however, requires the spacetime to be \textit{strongly hyperbolic}, meaning that the foliation $\mathcal{M}=\Sigma\times\mathbb{R}$ is allowed. 
This foliation can be understood as splitting the spacetime into a set of \textit{spacelike} hypersurfaces $\Sigma_t$. 


\paragraph{Spacelike Foliations}


Let the $t$ be the global smooth functions such that, 

\begin{equation}
\Sigma_{\tau} = \{x^{\alpha}\in\mathcal{M}: t(x^{\alpha})=\tau\},
\end{equation}

and let $\vec{t}$ be a vector such that $\langle\nabla t, \vec{t}\rangle = 1$. 
This $t$ can be seen as a "function that advances time" and $\vec{t}$ as a "flow of time" vector field.
Continuing the analogy, the rate at which a given tensor quantity $\boldsymbol{q}$ changes between hypersurfaces $\Sigma_t$ is given by the \textit{Lie derivative} of the $\boldsymbol{q}$ along the vector $\vec{t}$.

Consider two hypersurfaces $\Sigma_t$ and $\Sigma_{t+dt}$. 
A transition from one to another can be decomposed into the part tangent to the hypersurface $\Sigma_{t+dt}$ and expressed in a form of a vector $\vec{\beta}$ and a pert normal to the hypersurface $\Sigma_t$ and expressed as a $\alpha \vec{n}$, where $\vec{n}$ is a unit vector, normal to the $\Sigma_t$ in the diretion to $\Sigma_{t+dt}$. 
Then, the vector $\vec{t}$ can be written as 

\begin{equation}
\vec{t} = \alpha\vec{n}+\vec{\beta}.
\end{equation}

$\vec{\beta}$ is called \textit{shift vector} and $\alpha$ is called \textit{lapse-function}. 

The spacetime metric $\boldsymbol{g}$ can be decomposed into a spatial, \textit{Riemannian metric} $\boldsymbol{\gamma}$ as $\boldsymbol{\gamma} = \boldsymbol{g} + \underline{n} \otimes \underline{n}$, where $\underline{n}$ is the 1-form associated to the vector $\vec{n}$. \red{and?}
The \textit{Levi-Civita connection} can be computed by projecting the $\nabla$ on the space, tangent to the hypersurface $\Sigma_t$.

There are exist coordinates that are adapted to the foliation, namely $\{t, x^i\}$ with $\vec{\partial}_i\cdot \vec{n} = 0$. 
In these coordiantes the $\nabla t = dt$ and $\vec{t} = \vec{\partial}_t$. 

The connection between $\boldsymbol{g}$ and $\boldsymbol{\gamma}$ is $g_{\mu\nu}=\vec{\partial}_{\mu}\cdot\vec{\partial}_{\nu} $ and can be expressed in terms of $\alpha$ and $\vec{\beta}$ as

\begin{align}
\text{spatial components: } g_{ik}&=\vec{\partial}_{i}\cdot\vec{\partial}_{j} =\gamma_{ik}, \\
\text{time component: } g_{tt} &= \vec{\partial}_{t}\cdot\vec{\partial}_{t} = \vec{t}\cdot\vec{t} = - (\alpha^2-\vec{\beta}\cdot\vec{\beta}), \\
\text{mixed components: } g_{ti} &= \vec{\partial}_{t}\cdot\vec{\partial}_{i} = \vec{t}\cdot\vec{\partial}_i = (\alpha\vec{n}+\vec{\beta})\cdot\vec{\partial}_i=\beta_i,
\end{align}

where we made use of $\vec{\beta}$ being the spatial vector, \textit{i.e} $\vec{\beta}\cdot\vec{\beta}=\gamma_{ik}\beta^i\beta^k$.

The line-element can be thus written as
\begin{equation}
ds^2 = -(\alpha^2-\beta_i\beta^i)dt^2 +2\beta_i dx^i dt + \gamma_{ik} dx^i dx^k.
\end{equation}


\paragraph{Ex-curse: Hamiltonian Field Theory}


First we recall the generalized coordinates $\boldsymbol{q}$ and their \textit{covariant derivatives} $\nabla\boldsymbol{q}$. 

In light of the spacetime decomposition discussed above, we divide the $\boldsymbol{\alpha}$ into the time $dt$ and spatial parts represented by the \textit{antisymmetric symbol} ${^{(3)}\boldsymbol{\alpha}}$ as 

\begin{equation}
\boldsymbol{\alpha} = dx^0 \wedge dx^1 \wedge dx^2 \wedge dx^3 = dt \wedge {^{(3)}\boldsymbol{\alpha}}.
\end{equation}

Next, we introduce the "time derivative" as a \textit{Lie derivative along the vector field} $\vec{t}$ as 

\begin{equation}
\dot{\boldsymbol{q}} := \mathcal{L}_{\vec{t}}\boldsymbol{q}.
\end{equation}

As the $\Lambda(\boldsymbol{q}, \nabla\boldsymbol{q})$ is the \textit{Lagrangian density}, a conjugate momentum can be defined as 

\begin{equation}
\boldsymbol{p} := \frac{\partial\Lambda}{\partial\dot{\boldsymbol{q}}},
\end{equation}

Assuming that $\boldsymbol{p}$ and $\nabla\boldsymbol{q}$ can be expressed as a function of $\boldsymbol{q}$ and $\boldsymbol{p}$, inspired by the \textit{Legendre transformation}, we define the \textit{Hamiltonian} and its density as

\begin{align}
\mathcal{H} &= \boldsymbol{p}\cdot\dot{\boldsymbol{q}} - \mathcal{L}(\boldsymbol{q}, \nabla\boldsymbol{q}) \\
H &= \int_{\Sigma}\mathcal{H}{^{(3)}\boldsymbol{\alpha}}
\end{align}

Additionally we define the quantity 

\begin{equation}
J = \int_{0}^{t}H(\boldsymbol{q},\boldsymbol{p})dt = \int_{0}^{t}dt\int_{\Sigma}\mathcal{H}(\boldsymbol{q},\boldsymbol{p}){^{(3)}\boldsymbol{\alpha}} = \int_{0}^{t}dt\int_{\Sigma}{^{(3)}\boldsymbol{\alpha}}\Big(\boldsymbol{p}\cdot\dot{\boldsymbol{q}} - \mathcal{L}(\boldsymbol{q},\nabla\boldsymbol{q})\Big).
\end{equation}

Consider the variation of the $J$ with respect to the $\delta\boldsymbol{p}$ and $\delta\boldsymbol{q}$ as

\begin{equation}
\delta J = \int_{0}^{t}\delta H(\boldsymbol{q},\boldsymbol{p})dt = \int_{0}^{t}dt (\dot{\boldsymbol{q}}\delta\boldsymbol{p}+\boldsymbol{p}\delta\dot{\boldsymbol{q}}) - \int_{0}^{t}dt\delta\Lambda(\boldsymbol{q}, \nabla\boldsymbol{q}).
\end{equation}

Consider the last term. The variation of the Lagrangian is

\begin{equation}
\delta\Lambda = \int_{\Sigma}{^{(3)}\boldsymbol{\alpha}}\Bigg[\frac{\delta\Lambda}{\delta\dot{\boldsymbol{q}}}\delta\dot{\boldsymbol{q}}+\frac{\delta\Lambda}{\delta\boldsymbol{q}}\delta\boldsymbol{q}\Bigg],
\end{equation}

The first term in the square brackets can be reduced to $\boldsymbol{p}\delta\dot{\boldsymbol{q}}$, using the definition of the conjugate momentum. 
The second term can be treated, applying the Euler-Lagrange equations (\ref{eq:theory:eulerlagrange}). 
These manipulations result in

\begin{equation}
\delta\Lambda = \int_{0}^{t}dt\int_{\Sigma}{^{(3)}\boldsymbol{\alpha}}(\boldsymbol{p}\delta\dot{\boldsymbol{q}} + \dot{\boldsymbol{p}}\delta\boldsymbol{q}).
\end{equation}

Thus we obtain that 

\begin{equation}
\int_{0}^{t} \delta H(\boldsymbol{q},\boldsymbol{p})dt =   \int_{0}^{t}dt\int_{\Sigma}{^{(3)}\boldsymbol{\alpha}}(\dot{\boldsymbol{q}}\cdot\delta\boldsymbol{p}-\dot{\boldsymbol{p}}\cdot\delta\boldsymbol{q}),
\end{equation}

and as $\delta\boldsymbol{p}$ and $\delta\boldsymbol{p}$ are arbitrary, the \textit{Hamilton equations} read

\begin{equation}
\dot{\boldsymbol{q}}=\frac{\delta H}{\delta\boldsymbol{p}}, \hspace{5mm} \dot{\boldsymbol{p}} = -\frac{\delta H}{\delta\boldsymbol{q}}.
\label{eq:theory:hamiltoneqs}
\end{equation}

The Hamiltonian formalism can be used to re-derive the field-equations in a from that once the initial data is specified on a hypersurface $\Sigma_0$ for $\boldsymbol{q}$ and $\boldsymbol{p}$, the equations (\ref{eq:theory:hamiltoneqs}) would govern whole the evolution.


\paragraph{Extrinsic Curvature and Constraint equations}


We define the \textit{extrinsic curvature} of a $D-1$-surface $\Sigma_t\subset\mathcal{M}$ at a point $\mathcal{P}\in\Sigma_t$ as mapping $\boldsymbol{K}$ such that $\boldsymbol{K}(\boldsymbol{\upsilon})=-\nabla_{\boldsymbol{\upsilon}}\boldsymbol{n}$. 
Note, that the $\boldsymbol{K}$ does not depend on $\alpha$ and $\vec{\beta}$, it is a purely spatial tensor. 
The components of the extrinsic curvature are 

\begin{equation}
K_{\mu\nu} = -{\gamma^{\alpha}}_{\mu}\nabla_{\boldsymbol{u}}^{\alpha} n_{\nu} = -\frac{1}{2}\mathcal{L}_{\vec{n}}\gamma_{\mu\nu},
\label{eq:theory:extrcurvdef}
\end{equation}

where $\mathcal{L}_{\vec{n}}$ is the Lie derivative along the vector field $\vec{n}$. \\
From the (\ref{eq:theory:extrcurvdef}) the extrinsic curvature can be interpreted as a "speed of the $\vec{n}$ during the parallel transport along the hypersurface $\Sigma_t$".

\textit{Codazzi equations} relate the $4D$ Ricci tensor to the extrinsic curvature as

\begin{equation}
D_{\beta}K-D_{\alpha}{K^{\alpha}}_{\beta}=R_{\gamma\delta}n^{\delta}{\gamma^{\gamma}}_{\beta},
\label{eq:theory:formomentum}
\end{equation}

here $K$ is a trace of the tensor $\boldsymbol{K}$.

Gauss equation relates the $3D$ \textit{Riemann tensor} $^3{R_{\alpha\beta\gamma}}^{\delta}$ to the $4D$ one and the $\boldsymbol{K}$ as 

\begin{equation}
^3{R_{\alpha\beta\gamma}}^{\delta} = {\gamma^{\mu}}_{\alpha}{\gamma^{\nu}}_{\beta}{\gamma^{\lambda}}_{\gamma}{\gamma^{\delta}}_{\sigma}{R_{\mu\nu\lambda}}^{\delta}-K_{\alpha\gamma}{K_{\beta}}^{\delta}+K_{\beta\gamma}{K^{\delta}}_{\alpha}.
\label{eq:theory:forhamiltconst}
\end{equation}

The \textit{momentum constraint} thus can be obtained by substituting the (\ref{eq:theory:EFE}) into  (\ref{eq:theory:formomentum}) which yields

\begin{equation}
D_{\beta}K-D_{\alpha}{K^{\alpha}}_{\beta} = -8\pi{\gamma^{\alpha}}_{\beta} n^{\gamma}T_{\alpha\gamma}=:8\pi j_{\beta},
\label{eq:theory:momconstraint}
\end{equation}

where $j^{\alpha}$ is the ADM momentum density.

The \textit{Hamiltonian constraint} can be obtained by substituting EFE (\ref{eq:theory:EFE}) into the (\ref{eq:theory:forhamiltconst}), yielding 

\begin{equation}
^3 R+ K^2 - K_{\alpha\beta}K^{\alpha\beta} = 2G^{\alpha\beta}n_{\alpha}n_{\beta} = 16\pi n_{\alpha}n_{\beta} T^{\alpha\beta} =: 16\pi E,
\label{eq:theory:hamilconstraint}
\end{equation}

where $E$ is the ADM energy density.

The obtained constraint equations represent a set of \textit{elliptic equations} that must be satisfied on every hyprsurface $\Sigma_i$ of the foliation. 
It is however, possible to show that Einstein equations preserve the constraints, meaning that if they are satisfied at the initial slice $\Sigma_0$ they will be satisfied at any time in the future.


\paragraph{The Hamiltonian Formulation of the Einstein Equations, ADM equations}


\gray{deriving the evolution equations, ADM formulation}

%% Part 1 [Extrinsic curvature]

Here we briefly sketch to path of derivation of the Einstein field equations in the Hamiltonian framework. We will elude most of the intimidate and computationally extensive steps, as well as derivation of the boundary terms. For this we refer to \cite{Poisson:2004}.

First it is useful to note that determinant of the three-metric $\sqrt{\gamma}$ can be expressed as $\sqrt{\gamma}=\sqrt{-g}/\alpha$. 
The $p$ is the trace of the canonical momentum $\boldsymbol{p}$.

Now, consider the \textit{scalar curvature}, $R$,

\begin{align}
G_{\mu\nu} &= R_{\mu\nu} - \frac{1}{2}Rg_{\mu\nu} \\
-Rg_{\mu\nu}n^{\nu}n^{\mu} &= 2(G_{\mu\nu} n^{\nu}n^{\mu}-R_{\mu\nu}n^{\mu}n^{\mu})\\
-Rn_{\mu}n^{\mu}& = 2(G_{\mu\nu}n^{\nu}n^{\mu} - R_{\mu\nu}n^{\mu}n^{\mu}) \\
R &= 2(G_{\mu\nu}n^{\mu}n^{\nu} - R_{\mu\nu}n^{\mu}n^{\nu}).
\end{align}

From the \textit{Gauss-Codacci equation} (\ref{eq:theory:momconstraint}), which relates the spatial curvature $^{(3)}R$ to the spacetime curvature $R$, we have the following constraint relationship

\begin{equation}
2G_{\mu\nu}n^{\mu}n^{\nu} = {^{(3)}R} + K^2 - K_{\mu\nu}K^{\mu\nu}.
\end{equation}

The $R_{\mu\nu}n^{\mu}n^{\nu})$ can be expressed as a combination of extrinsic curvature and total divergences as follows.
From the definition of the Ricci tensor $R_{\mu\nu}$, we have:

\begin{align}
R_{\mu\nu} &= {R_{\mu\gamma\nu}}^{\gamma} \\
R_{\mu\nu}n^{\mu}n^{\nu} &= {R_{\mu\gamma\nu}}^{\gamma} \\
&= -(\nabla_{\mu}\nabla_{\gamma} - \nabla_{\gamma}\nabla_{\mu})n^{\gamma}n^{\nu} \\
&= n^{\mu}(\nabla_{\mu}\nabla_{\gamma} - \nabla_{\gamma}\nabla_{\nu})n^{\gamma} \\
&= (\nabla_{\mu}n^{\mu})(\nabla_{\gamma}n^{\gamma}) - \nabla_{\mu}(n^{\mu}\nabla_{\gamma}n^{\gamma}) - (\nabla_{\gamma}n^{\mu})(\nabla_{\mu}n^{\gamma}) + \nabla_{\gamma}(n^{\mu}\nabla_{\mu}n^{\gamma}) \\
&= K^2 - K_{\mu\gamma}K^{\mu\gamma} - \nabla_{\mu}(n^{\mu}\nabla_{\gamma}n^{\gamma}) + \nabla_{\gamma}(n^{\mu}\nabla_{\mu}n^{\gamma})
\end{align}

In case of variations with compact support, that we are interested in, the total divergences -- last two terms -- can be neglected. 
Then the result is

\begin{equation}
R_{\mu\nu}n^{\mu}n^{\nu}= K^2 - K_{\mu\nu}K^{\mu\nu}.
\label{eq:theory:rmunu_as_func_k}
\end{equation}

%% part 2 [Canonical momentum and Hamiltonian]

Using the fact that $\sqrt{\gamma}=\sqrt{-g}/\alpha$ and the (\ref{eq:theory:rmunu_as_func_k}) we obtain the Lagrangian density in terms of the variables on the hypersurface:

\begin{align}
\Lambda &= \sqrt{-g}R \\
&= \alpha\sqrt{\gamma}R \\
&= 2\alpha\sqrt{\gamma}(G_{\mu\nu}n^{\mu}n^{\nu} - R_{\mu\nu}n^{\mu}n^{\nu})\\ 
&= 2\alpha\sqrt{\gamma}\Big(\frac{1}{2}[{^{(3)}R} - K_{\mu\nu}K^{\mu\nu} + K^2] - K^2 - K_{\mu\nu}K^{\mu\nu}\Big)
\end{align}

Together with the contribution from matter fields, we obtain

\begin{equation}
\Lambda = \Lambda_g+\Lambda_m= \frac{1}{16\pi}\alpha({^{(3)}R} + K_{\mu\nu}K^{\mu\nu} - K^2)\sqrt{\gamma}+\Lambda_m
\end{equation}

Next we note that the extrinsic curvature of a
surface $\Sigma$ is defined as $K_{\mu\nu} = \nabla_{\mu}n_{\nu}$. 

To relate $K_{\mu\nu}$ to the metric, we make use of the following property of Lie derivatives:

\begin{align}
\mathcal{L}_{\vec{n}}g_{\mu\nu} &= n^{\gamma}\nabla_{\gamma}g_{\mu\nu} + g_{\gamma\nu}\nabla_{\mu}\upsilon^{\gamma} + g_{\mu\gamma}\nabla_{\nu}\upsilon^{\gamma} \\
&= \nabla_{\mu}n_{\nu}+\nabla_{\nu}\upsilon_{\nu} \\
&=2\nabla_{\mu}n_{\nu}
\end{align}

where the second line holds when $\nabla_{\gamma}\mu$ is the natural derivative operator corresponding to the metric $g_{\mu\nu}$ and the third line holds because $K_{\mu\nu}$ is symmetric.

Substituting this into our definition of $K_{\mu\nu}$,

\begin{align}
K_{\mu\nu} &= -\frac{1}{2}\mathcal{L}_{\vec{\vec{n}}}g_{\mu\nu} \\
&= -\frac{1}{2}\mathcal{L}_{\vec{\vec{n}}}(\gamma_{\mu\nu}-n_{\mu}n_{\nu}) \\
&= -\frac{1}{2}\mathcal{L}_{\vec{\vec{n}}}\gamma_{\mu\nu} \\
&= -\frac{1}{2}[n^{\gamma}\nabla_{\gamma}\gamma_{\mu\nu} + \gamma_{\gamma\nu}\nabla_{\mu}\upsilon^{\nu} + h_{\mu\gamma}\nabla_{\nu}\upsilon^{\gamma}] \\
&= -\frac{1}{2\alpha}[\alpha n^{\gamma}\nabla_{\gamma}\gamma_{\mu\nu} + \gamma_{\gamma\nu}\nabla_{\mu}\alpha\upsilon^{\nu} + h_{\mu\gamma}\nabla_{\nu}\alpha\upsilon^{\gamma}] \\
&= -\frac{1}{2\alpha}{\gamma_{\mu}}^{\gamma}{\gamma_{\nu}}^{\delta}[\mathcal{L}_{\vec{t}}\gamma_{\gamma\delta}-\mathcal{L}_{\vec{\beta}}\gamma_{\gamma\delta}] \\
&= -\frac{1}{2\alpha}{\gamma_{\mu}}^{\gamma}{\gamma_{\nu}}^{\delta}[\partial_t\gamma_{\mu\nu}-D_{\mu}\beta_{\nu}-D_{\nu}\beta_{\mu}]
\end{align}

and on the hypersurface $\Sigma$ the \textit{projection operators} are not needed. So we obtain

\begin{equation}
K_{\mu\nu} = -\frac{1}{2}\mathcal{L}_{\vec{n}}\gamma_{\mu\nu}=-\frac{1}{2\alpha}(\partial_t\gamma_{\mu\nu}-D_{\mu}\beta_{\nu}-D_{\nu}\beta_{\mu})
\end{equation}

which allows us to express the canonical momentum $p^{\mu\nu}$ as

\begin{align}
p^{\mu\nu} &= \frac{\partial\Lambda}{\partial\dot{\gamma}_{\mu\nu}} \\
&= -\frac{\sqrt{\gamma}}{16\pi}\alpha\Bigg[\frac{\partial {^{(3)}R}}{\partial\dot{\gamma}_{\mu\nu}} + \frac{\partial(K_{\mu\nu}K^{\mu\nu})}{\partial\dot{\gamma}_{\mu\nu}} - \frac{\partial K^2}{\partial\dot{\gamma}_{\mu\nu}}\Bigg] \\
&= \frac{\sqrt{\gamma}}{16\pi}(K\gamma^{\mu\nu} - K^{\mu\nu}),
\end{align}

where 

\begin{equation}
\frac{\partial K_{\mu\nu}}{\partial \dot{\gamma}_{\mu\nu}} = \frac{1}{2\alpha}, \hspace{5mm} \frac{\partial {^{(3)}R}}{\partial \dot{\gamma}_{\mu\nu}} = 0, \hspace{5mm}\frac{\partial K^2}{\partial \dot{\gamma}_{\mu\nu}} = \frac{\gamma^{\mu\nu}K}{\alpha}
\end{equation}

assuming that there is no explicit dependency of the $\Lambda$ on $dot{\gamma}_{\mu\nu}$.

Since, $\alpha$ and $\vec{\beta}$ are related to the the \textit{gauge freedom}, -- there are many ways the manifold $\mathcal{M}$ can be split into hypersurfaces -- the momenta associated with these functions and vectors is zero.

Thus, the \textit{Hamiltonian density} is

\begin{align}
\mathcal{H} &= p^{\mu\nu}\dot{\gamma}_{\mu\nu} - \Lambda \\
&= -\sqrt{\gamma}\alpha{^{(3)}R} + \frac{\alpha}{\sqrt{\gamma}}\Big[p^{\mu\nu}p_{\mu\nu}-\frac{1}{2}p^2\Big] + 2p^{\mu\nu} D_{\mu}\beta_{\mu} -\Lambda_m \\
%    &=  \frac{\sqrt{\gamma}}{16\pi}\Bigg\{\alpha\Big[-{^{(3)}R}+h^{-1}p^{\mu\nu}p_{\mu\nu}-\frac{1}{2}h^{-1}p^2\Big] - 2\beta_{\nu}\big[D_{\mu}(h^{-1/2}p^{\mu\nu})\big] + D_{\mu}(h^{-1/2}\beta_{\nu}p^{\mu\nu})\Bigg\} \\
&= \frac{\sqrt{\gamma}}{16\pi}\Bigg\{\alpha\Big[ -{^{(3)}R} + \gamma^{-1}p^{\mu\nu}p_{\mu\nu}-\frac{1}{2}\gamma^{-1}p^2\Big] +  2\beta_{\nu}\Big[D_{\mu}(\gamma^{-1/2}p^{\mu\nu})\Big] - 2D_{\mu}(\gamma^{-1/2}\beta_{\nu}p^{\mu\nu}) \Bigg\} - \Lambda_m,
\end{align}

where we restored the correct $16\pi$ factor in the last line.

As we consider variations with compact suppot, the last boundary term, can be neglected.

%% part 3 [Arriving at constraint equations (again)]

Now we consider the variation of the matter action $S_m$ with respect to the $\alpha$ and $\vec{\beta}$

\begin{align}
\frac{\delta S_m}{\delta \alpha} &=-\alpha\frac{\delta S_m}{\delta g_{00}} = -\alpha\sqrt{-g}T^{00} = -\alpha^2\sqrt{\gamma}T^{00} = -\sqrt{\gamma}T^{\mu\nu}n_{\mu}n_{\nu} \\
\frac{\delta S_m}{\delta \beta_{\mu}} &= \frac{\delta S_m}{\delta g_{\mu 0}} =\frac{1}{2}\sqrt{-g}T^{\mu 0} = -\frac{1}{2} \sqrt{\gamma}T^{\mu\nu}n_{\nu}.
\end{align}

As the variation of the Hamiltonian $H$ with respect to a quantity with vanishing canonical momentum is zero, we obtain two equations 

\begin{align}
\frac{\delta H}{\delta \alpha} &= 0 = -{^{(3)}R} + \gamma^{-1}p^{\mu\nu}p_{\mu\nu}-\frac{1}{2}\gamma^{-1}p^2 + 16\pi T^{\mu\nu}n_{\mu}n_{\nu} \\
\frac{\delta H}{\delta \beta_{\mu}} &= 0 = - D_{\mu}(\gamma^{-1/2}p^{\mu\nu}) + 8\pi{\gamma^{\mu}}_{\nu}n_{\gamma}T^{\nu\gamma}.
\label{eq:theory:hamiltonianvariation}
\end{align}

Note, that the $\delta H / \delta\beta_{\mu}$ is actually a \textit{Frech\'et differential} $dH$, $\delta \beta_{\mu}$, which is writes as

\begin{equation}
\langle dH,\delta\beta \rangle = \delta\beta_{\mu}\big[-D_{\nu}(\gamma^{-1/2}p^{\mu\nu})+8\pi n_{\gamma}T^{\mu\nu}\big], 
\end{equation}

containing $\delta\beta_{\mu}$ which is spatial. 

Thus only the spatial part is being constrained in the equation above. 
To account for that the projector ${\gamma^{\mu}}_{\nu}$ is added to the $\delta H/\delta \beta_{\mu}$.

The pair of equations (\ref{eq:theory:hamiltonianvariation}) is in fact the constraint equations derived before, namely the (\ref{eq:theory:momconstraint}) and (\ref{eq:theory:hamilconstraint}), and as we now see, they are related to the coordinate freedom of $\mathcal{M}$ decomposition and a coodrinate freedom on hypersurfaces.

%% Part 4 [hamiltonian equations for metric -- evolution equations]

Proceeding with the Hamiltonian formalism we note that equation \ref{eq:theory:hamiltoneqs} leads to the evolution equations for the three-metric, assuming that $\Lambda$ explicitly does not depend on the momentum

\begin{equation}
\dot{\gamma}_{\mu\nu} =\frac{\delta H}{\delta p^{\mu\nu}} = 2\gamma^{-1/2}\alpha\big(p_{\mu\nu}-\frac{1}{2}\gamma_{\mu\nu}p\big) - D_{\nu}\beta_{\mu}-D_{\mu}\beta_{\nu}
%    -2D_{(\mu}\beta_{\nu)},
\label{eq:theory:_adm_metric_evo}
\end{equation}

The evolution equations for the canonical momentum can read

\begin{align}
\dot{p}^{\mu\nu} = -\frac{\delta H}{\delta \gamma_{\mu\nu}} = &+ \alpha\gamma^{1/2}\big({^{(3)}R}^{\mu\nu}-\frac{1}{2}{^{(3)}R\gamma^{\mu\nu}}\big) \\
& - \frac{1}{2}\alpha\gamma^{-1/2}\gamma^{\mu\nu}\big(p_{\gamma\delta}p^{\gamma\delta}-\frac{1}{2}p^2\big) \\
& + 2\alpha\gamma^{-1/2}\big(p^{\mu\gamma}{p^{\nu}}_{\gamma}-\frac{1}{2}pp^{\mu\nu}\big) \\
& - \gamma^{1/2}\big(D^{\mu}D^{\nu}\alpha-\gamma^{\mu\nu}D^{\gamma}D_{\gamma}\alpha\big) \\
& - \gamma^{1/2}D_{\gamma}\big(\gamma^{-1/2}\beta^{\gamma}p^{\mu\nu}\big) \\
&+ 2p^{\gamma(\mu}D_{\gamma}\beta^{\nu)} + 8\pi \alpha \gamma^{1/2}S^{\mu\nu},
\label{eq:theory:_adm_mom_evo}
\end{align}

where $A_{(\mu\nu)} = 0.5(A_{\mu\nu}+A_{\nu\mu})$ the convention used, and where $S^{\mu\nu}={\gamma^{\mu}}_{\alpha}{\gamma^{\nu}}_{\beta}T^{\alpha\beta}$. 

Additionally, taking the variation of the matter field we noted that

\begin{equation}
\frac{\delta S}{\delta \gamma_{ik}} = \frac{\delta S_m}{\delta g_{ik}} = \frac{1}{2}\sqrt{-g}T^{ik}.
\end{equation}

The set of equations (\ref{eq:theory:hamiltonianvariation}), (\ref{eq:theory:_adm_metric_evo}) and (\ref{eq:theory:_adm_mom_evo}) comprise the ADM system. 
A more widely used from of these equations is in turns of $\gamma_{ij}$ and $K_{ij}$ that reads

\begin{align}
(\partial_t - \mathcal{L}_{\vec{\beta}})\gamma_{ik} &= -2\alpha K_{ik}; \\
(\partial_t - \mathcal{L}_{\vec{\beta}})K_{ik} &= -D_{i}D_{k}\alpha + \alpha\big(R_{ik} - 2K_{ij}{K^j}_k+KK_{ik}\big) - 8\pi\alpha\big(S_{ik} - \frac{1}{2}\gamma_{ik}(S-E)\big); \\
{^{(3)}R} + K^2 - K_{ik}K^{ik} &= 16\pi E; \\
D_{i}K-D_{k}{K^k}_i &= 8\pi j_i,
\label{eq:theory:adm}
\end{align}

where $S = \gamma^{ij}S_{ij}$.

These equations constitute the IVP for Einstein field equations and are known as ADM equations. 
The last two equations are the constraint equations. They determine how to set the initial data on the hypersurface $\Sigma_0$, via prescribing the three-metric and extrinsic curvature. The first two equations then govern the evolution.

\todo{make sure that the coefficients in formuals are consistent, $16\pi$ might me missing or $-$}
\todo{Makse sure that $\Lambda$ stands for largangian density and $\mathcal{L}$ for lie derivative}


\paragraph{Strongly Hyperbolic Formulations of the Einstein Equations}


It has been shown, that the ADM system of equations in its original form (\ref{eq:theory:adm}) is only \textit{weekly hyperbolic} \cite{Baumgarte:2002jm}. 
It was shown that in such system the errors tend to couple with zero-velocity modes \cite{Alcubierre:1999rt}. 

In an attempt to mitigate this problem, different formulations of the Einstein equations as initial-value problem were created. 
In particular, the \textit{generalized-harmonic formulation} \cite{Friedrich:1985,Lindblom:2005qh,Lindblom:2009}, 
the \textit{BSSNOK} formulation, derived by Baumgarte, Shapiro, Shibata, Nakamura, Oohara and Kojima \cite{Nakamura1987,Shibata:1995we,Baumgarte:1998te} 
the \textit{Z4} formulation \cite{Bona:2003fj,Bernuzzi:2009ex,Ruiz:2010qj,Weyhausen:2011cg,Alic:2011gg}. 
We do not attempt to elaborate on any of these formations and only aim to emphasize that a search for new and better formulations of Einstein equations for numerical applications is ongoing. 
We limit ourselves to sketching only the \textit{conformal-covariant variant of the Z4 formulation}, also known as \textit{Z4c}. 
The numerical implementation of this formulation was used to obtain the results discussed in this thesis. 


\paragraph{The CCZ4 Formulation}


The idea behind the Z4 formulation is to derive a set of evolution equations that is free from the zero-speed modes of the original ADM and thus -- \textit{strongly-hyperbolic}. 
This is achieved by not explicitly enforcing the constraints and treating the deviation from them as an dependent variable $Z_{\mu}$. The $Z_{\mu}$ is also called the Z4 four-vector.

First, consider the covariant Lagrangian

\begin{equation}
\Lambda = g^{\mu\nu}[R_{\mu\nu} + 2\nabla_{\mu}Z_{\nu}]\sqrt{g} + \Lambda_m,
\end{equation}

and applying \textit{Palatini-type variational principle} \cite{Bona:2010is}, leads to the evolution equations

\begin{equation}
R_{\mu\nu} + \nabla_{\mu}Z_{\nu} + \nabla_{\nu}Z_{\mu}=8\pi\Big(T_{\mu\nu} - \frac{1}{2}Tg_{\mu\nu}\Big),
\label{eq:theory:z4fieldeq}
\end{equation}

and two sets of constraint equations

\begin{equation}
\nabla_{\rho} g^{\mu\nu} = 0, 
\label{eq:theory:z4connect}
\end{equation}

and

\begin{equation}
Z_{\mu} = 0,
\end{equation}

where the latter is called an \textit{algebraic constraint}. 
If its derivative vanishes, it is equivalent to imposing the ADM momentum and Hamiltonian constraints \cite{Bona:2009}. 

The Einstein field equations themselves are recovered from (\ref{eq:theory:z4connect}) and (\ref{eq:theory:z4fieldeq}) when the algebraic constraint is satisfied. 

The Z4 system preserves the constraint, $\partial_t (Z_{\mu})= 0$. 
This allows to obtain the solution of the Einstein equations. 

However, the numerical solution of the system of equations introduces error, that leads to a constraint violation during the evolution. 
To mitigate this problem the Z4 system is further modified to enforce the dampening of the constraint violation propagation \cite{Gundlach:2005eh}.

A further modified version of Z4 was introduced by \cite{Alic:2011gg}. 
It incorporates the constraint-damping properties of the original Z4 and also allows for a better black hole treatment via \textit{moving-puncture}, that will be discussed later. 
The CCZ4 system reads 

\begin{align}
\partial_{t}\widetilde{\gamma}_{ij} = & -2\alpha\widetilde{A}_{ij}^{\text{TF}} + 2\widetilde{\gamma}_{k(i}\partial_{j)}\beta^k - \frac{2}{3}\widetilde{\gamma}_{ij}\partial_k \beta^k + \beta^k\partial_k\widetilde{\gamma}_{ij}, \\
\partial_{t}\widetilde{A}_{ij}^{\text{TF}} = & \phi^2\big[-\nabla_i\nabla_j\alpha + \alpha\big({^{(3)}R}_{ij}+\nabla_{i}Z_{j} + \nabla_{j}Z_{i}- 8\pi S_{ij}\big)\big]^{\text{TF}} \\
& + \alpha\widetilde{A}_{ij}(K-2\Theta)-2\alpha\widetilde{A}_{il}{\widetilde{A}^l}_{j} + 2\widetilde{A}_{k(i}\partial_{j)}\beta^{k} \\
& -\frac{2}{3}\widetilde{A}_{ij}\partial_{k}\beta^{k} + \beta^{k}\partial_{k}\widetilde{A}_{ij} \\
\partial_{t} \phi = & \frac{1}{3}\alpha\phi K - \frac{1}{3}\phi\partial_{k}\beta^{k} + \beta^{k}\partial_{k}\phi \\
\partial_{t}K = &-\nabla^{i}\nabla_{i}\alpha + \alpha\big({^{(3)}R} + 2\nabla_{i}Z^{i} + K^2 - 2\Theta K\big) + \beta^{j}\partial_{j}K \\
& - 3\alpha\kappa_1(1+\kappa_2)\Theta + 4\pi\alpha (S-3E) \\
\partial_{t}\Theta = &\frac{1}{2}\alpha\Big(R + 2\nabla_{i}Z^{i} - \widetilde{A}_{ij}\widetilde{A}^{ij} + \frac{2}{3}K^2 - 2\Theta K\Big) - Z^{i}\partial_{i}\alpha \\
& + \beta^{k}\partial_{k}\Theta - \alpha\kappa_1(2 + \kappa_2)\Theta - 8\pi\alpha E \\
\partial_{t}\hat{\Gamma}^j = & 2\alpha\Bigg({\widetilde{\Gamma}^i}_{jk}\widetilde{A}^{ij} - 	3\widetilde{A}^{ij}\frac{\partial_{j}\phi}{\phi} -\frac{2}{3}\widetilde{\gamma}^{ij}\partial_{j}K\Bigg) + 2\widetilde{\gamma}^{ki}\Big(\alpha\partial_{k}\Theta - \Theta\partial_{k}\alpha - \frac{2}{3}\alpha K Z_{k}\Big) \\
& - 2\widetilde{A}^{ij}\partial_{j}\alpha + \widetilde{\gamma}^{kl}\partial_{k}\partial_{l}\beta^{i} + \frac{1}{3} \widetilde{\gamma}^{ik}\partial_{k}\partial_{l}\beta^{l} + \frac{2}{3}\widetilde{\Gamma}^i\partial_{k}\beta^{k} \\
& - \widetilde{\Gamma}^k\partial_{k}\beta^{i} + 2\kappa_3\Big(\frac{2}{3}\widetilde{\gamma}^{ij}Z_{j}\partial_{k}\beta^{k} - \widetilde{\gamma}^{jk}Z_{j}\partial_{k}\beta^{i}\Big) + \beta^{k}\partial_{k}\hat{\Gamma}^i \\
& -2\alpha\kappa_1\widetilde{\gamma}^{ij}Z_{j}- 16\pi\alpha\widetilde{\gamma}^{ij}S_j,
\label{eq:theory:ccz4equations} % used for Whisky Code description
\end{align}

where $\Theta:=n_{\mu}Z^{\mu}=\alpha Z^0$, the $\widetilde{\Gamma}^i:=\widetilde{\gamma}^{jk}{\widetilde{\Gamma}^i}_{jk} = \widetilde{\gamma}^{ij}\widetilde{\gamma}^{kl}\partial_{l}\widetilde{\gamma}_{jk}$ and $\hat{\Gamma}:=\widetilde{\Gamma}^i + 2\widetilde{\gamma}^{ij}Z_j$, constants $\kappa_1$ and $\kappa_2$ are related to the constraint damping terms, the $\kappa_3$ is the additional constant for further adjustments, the three-dimensional Ricci tensor ${^{(3)})R}_{ij}$ is split into conformal part $\widetilde{R_{ij}^{\phi}}$ and the $\widetilde{R_{ij}}$ that contains the derivatives of the conformal metric

\begin{align}
\widetilde{R_{ij}} &= -\frac{1}{2}\widetilde{\gamma}^{lm}\partial_{l}\partial_{m}\widetilde{\gamma}_{ij} + \widetilde{\gamma}_{k(i}\partial_{j)}\widetilde{\Gamma}_{(ij)k} + \widetilde{\gamma}^{lm}\big[2\widetilde{\Gamma}^{k}_{l(i}\widetilde{\Gamma}_{j)km} + \widetilde{\Gamma}^{k}_{im}\widetilde{\Gamma}_{kjl}\big] \\
\widetilde{R_{ij}}^{\phi} &= \frac{1}{\phi^2}\big[\phi\big(\widetilde{\nabla}_{i}\widetilde{\nabla}_{j}\phi + \widetilde{\gamma}_{ij}\widetilde{\nabla}^{l}\phi\widetilde{\nabla}_{l}\phi\big) - 2\widetilde{\gamma}_{ij}\widetilde{\nabla}^{l}\phi\widetilde{\nabla}_{l}\phi\big]
\end{align}

And as one sees, the ecolution of $Z_i$ is now included in $\hat{\Gamma}$. 
\todo{understand the conformal stuff and add some steps to show how the ccz4 was made}


\paragraph{Gauge conditions}


During the discussion of the original ADM system, the choice of the lapse function, \textit{i.e} \textit{slicing condition}, and shift vector \textit{i.e} \textit{spatial gauge condition} was left open. 
The right choice however, is crucial for the stable evolution and in itself presents a broad and rapidly evolving subject. Here we are going to discuss only the gauge that is relevant for our work. 

\begin{itemize}
    \item \textit{Slicing conditions} 
    One of the widely used conditions is so called "maximal slicing" that sets $K=0$, which in turn results in the equation
    \begin{equation}
    D^{i}D_{i}\alpha = \alpha\big[K_{ij}K^{ij} + 4\pi(e+S)\big].
    \end{equation}
    This conditions has an advantage of being \textit{singularity-avoiding}. 
    For example, it was shown that in the case of Schwarzschild black hole, the $\alpha$ goes to zero at a finite distance from singularity \cite{Geyer:1995}. 
    However implementation of this condition in from of a \textit{elliptic equations} is computationally expensive.
    A class of slicing conditions in form of \textit{hyperbolic equations} that are more favorable from numerical standpoint and that reproduces the desired behavior of the maximal slicing was proposed in \cite{Bona:1994dr}. It is read 
    \begin{equation}
    (\partial_t - \beta^i\partial_i)\alpha = \alpha^2 f(\alpha)K
    \label{eq:theory:gauge_onepluslog}
    \end{equation}
    which in CCZ4 reads 
    \begin{equation}
    (\partial_t - \beta^i \partial_i )\alpha = \alpha^2 f(\alpha)(K-2\Theta)
    \end{equation}
    where $f(\alpha)$ is a positive function. 
    For many numerical applications, including those that are discussed in this work, the "$1 + \log$" slicing is adopted with the $\beta_i=0$. 
    Then, integrating equation (\ref{eq:theory:gauge_onepluslog}) yields 
    \begin{equation}
    \alpha = 1 + \log\gamma
    \end{equation}
    This condition is numerically more favorable and as $f\rightarrow\infty$ in the vicinity of a singularity, allows to treat black holes well like maximal slicing \cite{Baumgarte:2002jm}.
    \todo{add/modify some text.}
    
    \item \textit{Spatial gauge conditions}
    The requirements for the gauge are similar as in the case of the $\alpha$, namely hyperbolicity and minimization of numerical distortions for more stable evolution. 
    One of the widely used shift conditions is so called \textit{Gamma driver} condition \cite{Alcubierre:2002kk}, 
    \begin{align}
    \partial_t\beta^i &= \frac{3}{4}\alpha B^i, \\
    \partial_t B^i &= \partial_t\widetilde{\Gamma}^i - \eta B^i,
    \end{align}
    where $\eta$ is a dumping coefficient.
    This gauge condition tries to decrease the coordinate stretching that occur in the vicinity of a singularity. 
    It was shown to be effective in numerical applications, in particular for a single moving black hole. However it has a \textit{zero-speed mode}, that can amplify the numerical errors and destabilize the system \cite{vanMeter:2006vi}.
    A modified \textit{Gamma driver}, gauge that does not have zero or small speed modes:
    \begin{align}
    (\partial_t - \beta^j\partial_j)\beta^i &= \frac{3}{4}B^i \\
    (\partial_t - \beta^j\partial_j)B^i &= (\partial_t - \beta^j\partial_j)\widetilde{\Gamma}^i-\eta\beta^i,
    \end{align}
    was proposed by \cite{vanMeter:2006vi} and was applied to study binary black holes by \cite{Campanelli:2005dd}.
\end{itemize}


%% --- 

\subsection{Conclusion}

%% --- 

%% Subsection 1 [Euler-Lagrange equations]
In this section we have shown a derivation of \textit{Euler-Lagrange equations} from the variation of the action of \textit{Lagrange field theory} with respect to the \textit{generalized coordinates of the field}, and assuming the compact domain for boundary conditions.
%% Subsection 2 [The Hilbert Action]
Then, we sketched how the \textit{Einstein field equations} can be derived from the \textit{Einstein-Hilbert action}, that describes gravitational and matter fields, through the \textit{principle of least action}.
%% Subsection 3 [3+1 Decomposition of Einstein field equations]
Then we discussed how the Einstein field equations, which are the $10$ non-linear partial differential equations, can be decomposed into the $3+1$ system of equations, commonly adopted in numerical applications.
This is done by assuming the space-time to be strongly hyperbolic, and defining the initial data on the numm hypersurface and performing the space-time foliation, -- in other words -- splitting the spacetime into a set of \textit{spacelike} hypersurfaces.
There, the transition from one hyersurface to another is conveniently done with the help of vector $\vec{t} = \alpha\vec{n}+\vec{\beta}$. This "time flow vector" is composed of the part tangent to the hypersurface (to which the transition is done), the so-called \textit{shift}, $\vec{\beta}$; and the part, normal to the previous hypersurface in the direction of the next one, the lapse function $\alpha$. 
The space-time metric is then decomposed into the spatian, \textit{Riemannian metric} and time part.
%% paragraph []
In light of the spacetime decomposition, the Hamiltonian formalism can be used to re-derive the field-equations in a from that once the initial data is specified on a hypersurface $\Sigma_0$ for $\boldsymbol{q}$ and $\boldsymbol{p}$, the equations (\ref{eq:theory:hamiltoneqs}) would govern whole the evolution. 
Such form consists of a set of constraint and evolution equations.
%% Extrisnsic curvature and momentum constraint.
To obtain them, first the concept of the extrinsic curvature is introduced, which can be thought of as a speed of the norm to the hypersurface, $\vec{n}$, during its parallel transport along the hypersurface $(\Sigma_t)$.
%% constraint equations
Then, the first, \textit{momentum constraint}, can be obtained by inserting the EFE intor the Codazzi equation, that relates the $3D$ Ricci tensor to the extrinsic curvature. 
The second, the \textit{Hamiltonian constraint} is derived by substituting the EFE into the gauss equation that relates the $4D$ Riemann tensor with the $3D$ one and extrinsic curvature.
%% on the constraint equations
The obtained constraint equations represent a set of \textit{elliptic equations} that must be satisfied on every hyprsurface $\Sigma_i$ of the foliation. 
It is however, possible to show that Einstein equations preserve the constraints, meaning that if they are satisfied at the initial slice $\Sigma_0$ they will be satisfied at any time in the future. 
%% paragraph [The Hamiltonian Formulation of the Einstein Equations]
The derivation of the evolution equations can be summarized as follows. 
%% --- Part 1 [Extrinsic curvature]
We begin by expressing the extrinsic curvature $R$, through $G_{\mu\nu}$ and $R_{\mu\nu}$. 
Then, we utilize \textit{Gauss-Codacci equation} (the momentum constraint), that relates spatial curvature $^{(3)}R$ to the spacetime curvature $R$.
Next, we note that from its definition, the Ricci tensor, $R_{\mu\nu}$, can be expressed as a combination of extrinsic curvature, $K$, and $K_{\mu\gamma}K^{\mu\gamma}$ and total divergences, which in the case of variations with compact support, nullifies. 
The result is a simple relation between the Ricci tensor $R_{\mu\nu}$ and extrinsic curvature $K$ and $K_{\mu\nu}K^{\mu\nu}$, written as 
$R_{\mu\nu}n^{\mu}n^{\nu}= K^2 - K_{\mu\nu}K^{\mu\nu}$
%% --- Part 2 [Deriving Canonical momentum and Hamiltonian]
Now we can write the Lagrangian density, $\Lambda$ in terms of the variables on the hypersurface, which consists of the gravity and matter parts, as $\Lambda = \Lambda_g+\Lambda_m$. 
Then, we relate the extrinsic curvature on the hypersurface, $K_{\mu\nu}$, to the metric, making use of a certain property of Lie derivative. 
Then we arrive at an expression for the canonical momentum, $p^{\mu\nu}$, and in turn, to the Hamiltonian density $\mathcal{H}$.
%% --- Part 3 [Arriving at constraint equations (again)]
Now we consider the variation of the matter action $S_m$ with respect to the $\alpha$ and $\vec{\beta}$ and the variation of the Hamiltonian $H$.
This gives a set of equations, in which only the spatial part is constrained, which requires an addition of a projector. Thus, we have arrived to the pair of constrained equations, derived before. However, now we see that they are related to the coordinate freedom of $\mathcal{M}$ decomposition and a coordinate freedom on hypersurfaces.
%% --- Part 4 [hamiltonian equations for metric -- evolution equations]
Recalling the Hamiltonian system of equations (canonical equations), we note that they lead to the evolution equations for the three-metric, $\dot{\gamma}_{\mu\nu}$. Then we obtain the evolution equations for the canonical momentum $\dot{p}^{\mu\nu}$. The set of equations we obtain, then, comprise the so-called ADM system.
%% --- Remarks 
These equations constitute the IVP for Einstein field equations and are known as ADM equations. The last two equations are the constraint equations. They determine how to set the initial data on the hypersurface $\Sigma_0$, via prescribing the three-metric and extrinsic curvature. The first two equations then govern the evolution.
%% Hyperbolicity -- Formulations
The ADM formulation in its basic form, derived as described above, is however weakly hyperbolic \cite{Baumgarte:2002jm} and not very well suited for numerical applications, as errors tend to couple to non-zero velocity modes \cite{Alcubierre:1999rt}.
A search for a better formulation of EFE is an ongoing effort. Following formualtions have been developed. 
The \textit{generalized-harmonic formulation} \cite{Friedrich:1985,Lindblom:2005qh,Lindblom:2009}, 
the \textit{BSSNOK} formulation, derived by Baumgarte, Shapiro, Shibata, Nakamura, Oohara and Kojima \cite{Nakamura1987,Shibata:1995we,Baumgarte:1998te} 
the \textit{Z4} formulation \cite{Bona:2003fj,Bernuzzi:2009ex,Ruiz:2010qj,Weyhausen:2011cg,Alic:2011gg}.
In the following we briefly discuss the formulation of relevance to this thesis, the \textit{conformal-covariant variant of the Z4 formulation}, also known as \textit{Z4c}.
%% --- CCZ4
The idea behind the Z4 formulation is to derive a set of evolution equations that is free from the zero-speed modes of the original ADM and thus -- strongly-hyperbolic. 
This is achieved by not explicitly enforcing the constraints and treating the deviation from them as an dependent variable $Z_{\mu}$. The $Z_{\mu}$ is also called the Z4 four-vector.
%% --- derivation 
The procedure involves invoking again the Lagrangian, (ogf gravity and matter) and applying \textit{Palatini-type variational principle} \cite{Bona:2010is}. This results in a evolution equations and two sets of constraint equations and $Z_{\mu}=0$ equation, which is called the algebraic constraint. If its derivative vanishes, it is equivalent to imposing the ADM momentum and Hamiltonian constraints \cite{Bona:2009}. 
Then, the Einstein field equations themselves are recovered from the obtained set of evolution and constraint equations if the algebraic constraint is satisfied.
The Z4 system preserves the constraint, $\partial_t (Z_{\mu})= 0$. This allows to obtain the solution of the Einstein equations. 
However, the numerical solution of the system of equations introduces error, that leads to a constraint violation during the evolution. To mitigate this problem the Z4 system is further modified to enforce the dampening of the constraint violation propagation \cite{Gundlach:2005eh}.
A further modified version that incorporates the constraint-damping properties of the original Z4 and also allows for a better black hole treatment via \textit{moving-puncture} was introduced by \cite{Alic:2011gg}.
%% --- Gauge conditions
During the derivation of the ADM system, the choice of the lapse function, \textit{i.e} slicing condition, and shift vector \textit{i.e} spatial gauge condition was left open. The right choice however, is crutual for the stable evolution and in itself presents a broad and rapidly evolving subject.
The \textit{Slicing condition} called 'maximal slicing', for instance, has a property of being \textit{singularity-avoiding}, -- avoiding $\alpha$ going to $\infty$ at the vicinity of a singularity of a Schwarzschild black hole \cite{Geyer:1995}.
Of particular relevance is the so-called "$1 + \log$" slicing, that states th.at $\beta_i=0$ and $\alpha = 1 + \log\gamma$, where $\gamma$ is \gray{trace of the three-metric}. This condition is numerically more favorable and as $f\rightarrow\infty$ in the vicinity of a singularity, allows to treat black holes well like maximal slicing \cite{Baumgarte:2002jm}.
The requirements for the \textit{Spatial gauge conditions} are similar as in the case of the $\alpha$, namely hyperbolicity and minimization of numerical distortions for more stable evolution. 
One of the widely used shift conditions is so called \textit{Gamma driver} condition \cite{Alcubierre:2002kk}. It has however some undesired numerical properties -- zero-speed mode, that can amplify the numerical errors and destabilize the system \cite{vanMeter:2006vi}. These problems are addressed by the modified \textit{Gamma driver}, gauge that does not have zero or small speed modes that was proposed by \cite{vanMeter:2006vi} and applied to study binary black holes by \cite{Campanelli:2005dd}.