%% =====================================================================================
%%
%%               P O S T - P R O C E S S I N G
%%
%% =====================================================================================
\chapter{Postprocessing methods} \label{app:ppr}

%% =====================================================================================
%%
%%               P O S T - P R O C E S S I N G
%%
%% =====================================================================================


%\section{Postprocessing tools and methods}

In this appendix we discuss the methods used to analyze the simulation data,
focusing on the overall average quentities that can be used for comparisons 
of the remnant, disk and ejecta. 
%
For this we mainly use the mass-averaged quantities defined as 
%
\begin{equation}
\langle f \rangle = \frac{\sum_i f(m_i)m_i}{\sum_i m_i}
\end{equation}
%
where $m_i$ is the mass contained in the $i$-th bin.



\section{Disk \& Remnant}\label{sec:bns_sims:method:disk}

It is common to discuss the \pmerg{} state of the \ac{BNS} systems in terms of the remnant, 
a \ac{MNS} or a \ac{BH}, and a disk or torus surrounding it. 
%However there is not unified convention in how to define the disk and separate it from the former. 

A disk around a \ac{BH} is the matter outside the apparent horizon 
\citep[\eg]{Dietrich:2015iva,Dietrich:2016hky}. 
%However, owing to the disk accretion onto a \ac{BH}, the evaluation time is a 
%crucial parameter, and unfortunately is not consistent in the literature, 
%(\eg $\sim1$~ms in \citet{Dietrich:2015iva,Dietrich:2016hky} and $\sim30$~ms in \citet{Sekiguchi:2016bjd}).

A disk around a \ac{MNS} is more difficult to define. Commonly, the 
threshold $\rho\sim 10^{13}$~\gcm is assumed. The choice is motivated by studies that 
showed that the angular velocity profiles becomes approximately Keplerian at this point 
\citep[\eg][]{Shibata:2005ss,Shibata:2006nm,Hanauske:2016gia,Kastaun:2016elu}.
This convection was adopted in several recent works 
\citep{Radice:2018pdn,Kiuchi:2019lls,Vincent:2019kor}.
The extraction time here is also important, but less so, as we find that the accretion on the 
\ac{NS} is considerably slower.
%In the case where the remnant is a neutron star, the disk definition usually includes 
%the density cut. For instance, in \cite{Radice:2018pdn,Kiuchi:2019lls,Vincent:2019kor} 
%the disk is assumed to encompass the matter with $\rho < 10^{13}$~\gcm. 
%The threshold $\rho\sim 10^{13}$~\gcm corresponds to the point in the remnant where
%the angular velocity profiles becomes approximately Keplerian, 
%\citep[\eg][]{Shibata:2005ss,Shibata:2006nm,Hanauske:2016gia,Kastaun:2016elu}.
%The extraction time here is also important, but less so, as we find that the accretion 
%on the NS is considerably slower.
%
%Overall, we estimate that these differences can amount to a systematic factor of a few,
%which we employ for the statistical analysis in section \ref{sec:stat:anal}

% In this thesis we define the disk as a matter that satisfies two criteria 
%$\alpha > 0.15$ and $\tho < 10^{13}\gcm$, where $\alpha$ is the lapse function 
%(see section \ref{sec:theory:gr3p1}).

We compute the disk baryonic mass of the disks evaluating
the volume integral of the conserved rest-mass density $D=\sqrt{\gamma}~W\rho$,
%
\begin{equation}
\label{eq:method:mdisk}
M_{\text{disk}} = \int D \dd^3 x
\end{equation}
%
from $3$D snapshots of the simulations.% in postprocessing.

%% In \cite{Sekiguchi:2016bjd}, the disk mass is extracted at
%% ${\approx} 30$~ms outside the AH. In \cite{Radice:2018pdn}, the disk mass is computed
%% as the baryonic mass outside the AH at BH formation, while for NS
%% remnants the criterion $\rho < 10^{13}$ g cm$^{-3}$ is used. 
%% In \cite{Kiuchi:2019lls} for both BH and NS outcome the $\rho < 10^{13}$ g cm$^{-3}$ 
%% criterion is used and time of the extraction is not specified. 
%% In \cite{Vincent:2019kor} the density criterion is the same, however the simulations 
%% are significantly shorter (${~\sim 7.5}$~ms) than in other
%% works. Overall, we estimate that these differences can amount to a
%% systematic factor of a few.



\section{Density modes} \label{sec:bns_sims:method:modes}

%% FROM THE LETTER 

The hydrodynamic instability in newly formed \acp{MNS} 
is monitored by a decomposition in Fourier modes
$e^{- i m \phi}$ of the Eulerian rest-mass density on the equatorial plane 
[see Eq.~(1) of \citep{Radice:2016gym}] and characterized by the
development of a $m=2$ followed by a $m=1$ mode 
\citep{East:2015vix,Paschalidis:2015mla,Radice:2016gym,Lehner:2016wjg,Bernuzzi:2013rza,Kastaun:2014fna}.
%In the short-lived remnant (LS220) the $m=1$ mode
%is subdominant with respect to the $m=2$, and it reaches a maximum close to the collapse
%\citep{Bernuzzi:2013rza}. Instead, in the long-lived remnant (DD2) the $m=1$
%becomes the dominant mode at $\sim$20~ms and persists throughout the
%remnant's lifetime, while the $m=2$ efficiently dissipates via
%gravitational-wave emission \citep{Bernuzzi:2015opx,Radice:2016gym}.
%
%% FROM THE PAPER 
%
%During the post-merger evolution the neutron star oscillates the most prominant modes 
%are quasiradial mode $F$ ($m=0$), the $m=2$ $f$-model 
%and non-linear combinations of them \citep[\eg][]{Shibata:2000jt,Stergioulas:2011gd}.
%
%It has also been shown that the $m=1$, so-called one-armed spiral instability, 
%is present in the remnant of \ac{BNS} mergers  
%\citep{Paschalidis:2015mla,Radice:2016gym,East:2016zvv}.
%
%In order to investigate the dynamical instabilities in our simulations we 
%project the rest-mass density onto spherical harmonics,
%or, in other words, we perform the complex azimuthal mode decomposition of,
%the conserved rest-mass density.
For the data availability reasons we consider only $\rho(x,y,z=0,t)$, 
\ie restrict our analysis to the orbital plane $z=0$

%\red{DOUBLE check the presence of Gamma and W}
%% \begin{equation}
%% \label{eq:modes}
%% C_m = \int \rho(x,y,z=0,t) W e^{-i m \phi} \sqrt{\gamma} %% \text{d}x \text{d} y \, ,
%% \end{equation}
\begin{equation}
\label{eq:modes}
C_m(t) = \int \rho(x,y,z=0,t) e^{-i m \phi(x,y)} \text{d}x \text{d} y \, ,
\end{equation}
%(see \eg~\citet{Baiotti:2009gk}).

% where %% $\rho$ is the rest mass density, 
% $\gamma$ is the determinant of the three-metric and $W$ is the
% Lorentz factor between the fluid and the Eulerian observers. 
%
We note that the above quantities are gauge dependent.
%
%\red{Dietrich in his thesis thinks that the growing m=1 mode is "We can not
%    exclude the possibility that the growing m = 1 mode is triggered by numerical effects,
%    but we think that it is a physical hydrodynamical effect due to mode couplings."
%    Page 53 of the thesis
%}
%
In addition, frequencies of the modes can be computed with the Fourier analysis 
of the $\rho_{\text{max}}$ and projections $C_{m}$. We resort it for the future work, 
as in this work we are interested only in their magnitude. 





\section{Angular momentum} \label{sec:bns_sims:method:ang_mom}

%\red{First, white that in GR the angular momentum is not clearly defined}

%% FROM LETTER 

%From the fluid's stress energy tensor,
%we compute the angular momentum density flux $J_r = T_{ra}(\partial_\phi)^a$,
%where $\phi$ is the cylindrical angular coordinate;
%angular momentum is conserved if $(\partial_\phi)^a$ is a Killing vector.
%
%% FROM PAPER 

The fluid's angular momentum analysis in the remnant and disk is performed
assuming axisymmetry,
%(see Appendix~\ref{app:ang} for derivation)
that is, we assume $\phi^{\mu} = (\partial_{\phi})^{\mu}$ is a Killing
vector. Accordingly, the conservation law (Eq.~\eqref{eq:theory:tmunu_eq_0}) 
reads
%
\begin{equation}
\partial_t(T^{\mu\nu}\phi_{\nu}n_{\nu}\sqrt{\gamma}) -
\partial_i(\alpha T^{i \nu}\phi_{\nu}\sqrt{\gamma}) = 0 \ ,
\end{equation}
%
where $n^\mu$ is the normal vector to the spacelike hypersurfaces of
the spacetime's $3+1$ decomposition.
%
The equation implies the conservation of the angular momentum 
\begin{equation}
J = % \int j dV = 
%- \int \, T_{\mu\nu}n^{\mu}\phi^{\nu}\,\dd ^3x = 
-\int \,
T_{\mu\nu}n^{\mu}\phi^{\nu}\,\sqrt{\gamma}\, \dd^3 x\ .
\end{equation}
%
In the cylindrical coordinates $x^i=(r,\phi,z)$ adapted to the symmetry
the angular momentum density is  
%
\begin{equation}
j = %-
\rho h W^2 v_{\phi} \ ,
\label{eq:method:ang_mom}
\end{equation}
%
and the angular momentum flux is 
%
\begin{equation}
\alpha\sqrt{\gamma}T^r _{\nu}\phi^{\nu} =
\alpha\sqrt{\gamma}\rho h W^2 (v^{r}v_{\phi}) .
\end{equation}
%
We evaluate these quantities from the 3D snapshots of our simulations.




\section{Ejecta} \label{sec:bns_sims:method:ejecta}

To model and study the electromagnetic counterparts to mergers, the amount and properties 
of the material leaving the system are needed.
The matter expelled at high velocity may ultimately become unbound from the 
central gravitational potential. 
There are two indicators commonly adopted to mark the unbound matter.

\subsection{The Geodesic criterion}

Assuming that the spacetime is stationary, ($\partial_t$ is the Killing vector)% \red{confirm}, 
the four-velocity, $u_t$, (along the time-like killing vector), 
is a constant of motion for geodesics. 
Additionally, if the space is asymptotically flat, at infinity the $u_t = -W$, 
where $W$ is the fluid element Lorentz factor. 
Then, if a fluid element has $u_t < -1$, it may be considered unbound 
as it will retain the non-zero positive velocity at infinity. 
%
The fluid asymptotic velocity 
\begin{equation}
\upsilon_{\infty} \simeq \sqrt{2E_{\infty}} = \sqrt{(1-u_t ^2)}.
\end{equation}
%
This criterion can be through of as considering the fluid to be made of isolated 
particles that follow the geodesics. Indeed, the effects of \acp{EOS}, 
fluids pressure gradient, internal energy and heating %(\eg, due to an $r$-process (\red{see section XXX})) 
are neglected which might lead to 
underestimation of the ejecta mass. 
%
The space time is also assumed to be static.
%
Strictly speaking, none of these assumptions is 
fulfilled in the \ac{BNS} post-merger environment. However, this criterion 
is widely used in the literature \citep[\eg][]{Radice:2018pdn,Vincent:2019kor}.



\subsection{The Bernoulli criterion}

From the relativistic Bernoulli equation \citep{Rezzolla:2013}, it follows
that for a stationary relativistic flow, the $hu_t$ is constant along the 
fluid worldliens. Here $h$ is the (relativistic) enthalpy, which is
defined up to a constant factor. 
%
If at the spatial infinity the enthalpy is set so $h\rightarrow-1$, 
%\red{in Vincent it is $h\leftarrow 1$}
the condition $hu_t < -1$ would mark the unbound matter 
(as in the assymptotically flat space-time the 
$u_t = -W$ for the flow particles following geodesics).
%
The associated asymptotic velocity is calculated as 
%
\begin{equation}
\upsilon_{\infty} \simeq \sqrt{2 (h (E_{\infty}+1)-1)}. 
\end{equation}
%
The criterion can be regarded as assuming all the internal energy of the fluid 
gets added to the fluid kinetic energy, as the fluid decompresses \red{(pressure drops?)}.
%
%The $r$-process nucleosynthesis that occurs in the outflow deposits the energy.
%\red{In Vincent it is assumed that the difference in binding energy between the 
%    particles in NSE at a given $\rho$, $T$ and $Y_e$, and their binding energy at
%    the same $Y_e$ but low $\rho$ and $T$ is added/substracted from the fluid's kinetic energy.
%    Out-of-NSE evolution and effects this neglected \citep[][see]{Foucart:2016vxd}
%    [BUT I AM NOT SURE IF THIS IS THE CASE FOR OUR SIMULATIONS! MAYBE NOT!]}
%
This criterion has been reported to estimate the amount of unbound material more accurately
\citep{Foucart:2015gaa}. 
Moreover, it was reported that the Bernoulli criterion leads to up to twice 
the amount of ejecta detected in comparison with the geodesic criterion, 
if the estimation is done within a given volume \citep{Kastaun:2014fna}.



We adopt the geodesic criterion to study the "burst-like", short outflows,
such as dynamical ejecta, where the pressure gradient is not expected to make a significant contribution.
For the steady-state outflows, like \pmerg{} winds we adopt the Bernoulli criterion.
%
The term ejecta would refer to the material gravitationaly 
unbound according to either of the criteria.
%
All ejecta properties are evaluated at $R \simeq 294$ from the center of the 
simulation domain unless stated otherwise. 