% Appendix Template

\chapter{Appendix Title Here} % Main appendix title

\label{app:fit} % Change X to a consecutive letter; for referencing this appendix elsewhere, use \ref{AppendixX}

This appendix summarizes all fit coefficients.
Dynamical ejecta coefficients can be found in 
Tab.~\ref{tab:dynfit:poly} and 
Tab.~\ref{tab:dynfit:fit_form} for the polynomials and fitting
formulae respectively.
Disk coefficients can be found in 
Tab~\ref{tab:diskfit:poly} and 
Tab.~\ref{tab:diskfit:form} for the polynomials and fitting
formulae respectively.
The coefficients of the recommended fitting formulae, as discussed in
the conclusion, are highlighted in the tables.

%% --------------------------------
\input{Tables/tabFits.tex}  
%% Coefficients
%% --------------------------------



\red{NOT YET APPROVED}
\red{NOT rephrased }
\section{$\chi_\nu^2$ statistics for individual datsets}
\label{app:datasets}

In this appendix we discuss the $\chi_\nu^2$ statistics of all fitting formulae 
dataset-vise instead of adding them up, as was done in the main text.
%% --- 
In Tab.~\ref{tbl:fit:ejecta:chi2dofsallInd} we compare the different fits 
for the dynamical ejecta properties and disk mass in terms of the reduced 
$\chi$-squared, $\chi^2 _{\nu}$.
%% --- ejecta mass
Regarding the ejecta mass we find that the datasets that are 
more uniform in their physics and method, \eg, \DSrefset{} and 
\DScool{} display the lowest $\chi^2 _{\nu}$.
The largest $\chi^2 _{\nu}$ is found for the \DSheatcool{}, that 
incorporates both, models with M1 and leakage+M0 neutrino schemes. 
This further highlights the systematic uncertainties due 
to different neutrino treatment. 
%% --- ejecta velocity 
For the ejecta velocity, the largest $\chi^2 _{\nu}$ is found 
for the \DScool{} across all fitting formulae. 
This might be attributed to the systematic uncertainties that pure 
leakage neutrino scheme introduces for models with different outcomes, \eg,
prompt collapse and stable remnants. 
%% --- ejecta electron fraction and RMS angle
With respect to ejecta electron fraction and RMS half-opening angle, 
the difference in $\chi^2 _{\nu}$ are small. Notably, for the 
$\ayd$, the $\chi^2 _{\nu}$ is similar between the \DSrefset{}
and \DSheatcool{}. This indicates the consistency in neutrino reabsorption 
effects on the ejecta composition in these datasets.
%% --- disk mass
For the disk mass we find the largest $\chi^2 _{\nu}$ for \DSrefset{} 
and \DScool{} across all fitting formulae. 
This can be explained by the presence of prompt collapse simulations 
in both datasets.
However, a larger set of models cover the binary parameter space more 
uniformly is required to investigate this further.
%% --- conclusion
%% Overall we find that for all quantities the $P_2^2{q,\tilde{\Lambda}}$ is 
%% a reasonably good fitting model. 





%% --------------------------------
\input{tabRedChi2IndDat.tex}  %% tbl:fit:ejecta:chi2dofsallInd
%% Coefficients
%% --------------------------------
